Diagnostics & Interactive Dynamics of the Colorectal Cancer Virome

### 2017-07-11 doc/manuscript.Rmd
Updated man and cl.

---
title: Diagnostic Potential & Interactive Dynamics of Viral & Bacterial Communities in Colorectal Cancer
author:
- name: Geoffrey D Hannigan
  affiliation: 1
- name: Melissa B Duhaime
  affiliation: 2
- name: Mack T Ruffin IV
  affiliation: 3
- name: Charlie C Koumpouras
  affiliation: 1
- name: Patrick D Schloss
  affiliation: 1,\*
address:
  - code: 1
    address: Department of Microbiology & Immunology, University of Michigan, Ann Arbor, Michigan, 48109
  - code: 2
    address: Department of Ecology & Evolutionary Biology, University of Michigan, Ann Arbor, Michigan, 48109
  - code: 3
    address: Department of Family & Community Medicine, Pennsylvania State University Hershey Medical Center, Hershey, Pennsylvania, 17033
  - code: \*
    address: To whom correspondence may be addressed.
output:
  md_document:
    variant: markdown
abstract: Colorectal cancer is a primary cause of morbidity and mortality throughout the world. Colorectal cancer development has been linked to differences in colonic bacterial community composition, however viruses have yet to be studied in colorectal cancer despite their oncogenic potential. We used 16S rRNA gene, whole shotgun metagenomic, and purified virus metagenomic sequencing of stool to evaluate the differences in human colorectal cancer virus and bacterial community composition. Through random forest modeling we identified differences in the healthy and colorectal cancer virome. The cancer-associated virome consisted primarily of temperate bacteriophages that were also bacteria-virus community network hubs. These results provide foundational evidence that bacteriophage communities are associated with colorectal cancer and likely impact cancer progression by altering the bacterial host communities.
font: Times
fontsize: 12pt
---

@@

## Cohort Design, Sample Collection, and Processing
Our study cohort consisted of 90 human subjects, 30 of whom had healthy colons, 30 of whom had adenomas, and 30 of whom had carcinomas **(Figure \ref{sampleproc})**. Half of each stool sample was used to sequence the bacterial communities using both 16S rRNA gene and shotgun sequencing techniques. The 16S rRNA gene sequencing was performed for a previous study, and the sequences were re-analyzed using contemporary methods [@Zackular:2014fba]. The other half of each stool sample was purified for virus like particles (VLPs) before genomic DNA extraction and shotgun metagenomic sequencing. In the VLP purification, cells were disrupted and extracellular DNA degraded **(Figure \ref{sampleproc})** to allow the exclusive analysis of viral DNA within virus capsids. In this manner, the *extracellular virome* of encapsulated viruses was targeted.

@@

## Unaltered Virome Diversity in Colorectal Cancer
Microbiome and disease associations are often described as being of an altered diversity (i.e., "dysbiotic"). Therefore, we first evaluated the influence of colorectal cancer on virome OVU diversity. We evaluated differences in communities between disease states using the Shannon diversity, richness, and Bray-Curtis metrics. We observed no significant alterations in either Shannon diversity or richness in the diseased states as compared to the healthy state **(Figure \ref{betaogu} C-D)**. There was no statistically significant clustering of the disease groups (ANOSIM p-value `r pvalcalc(alphadf[c(alphadf$Title %in% "anovawithoutneg"),"Stat"])`, **Figure \ref{betaogu}**). Notably, there was a significant difference between the few blank controls that remained after rarefying the data and the other study groups (ANOSIM p-value `r pvalcalc(alphadf[c(alphadf$Title %in% "anovawithneg"),"Stat"])`, **Figure \ref{betaogunegative})**, further supporting the quality of the sample set. In summary, standard alpha and beta diversity metrics were insufficient for capturing virus community differences between disease states **(Figure \ref{betaogu})**. This is consistent with what has been observed when the same metrics were applied to 16S rRNA sequenced and metagenomic samples [@Zeller:2014ix; @Zackular:2014fba; @Baxter:2016dja] and points to the need for alternate approaches to detect the impact of colorectal cancer disease state on these communities.

## Altered Virome Composition in Colorectal Cancer
As opposed to the diversity metrics discussed above, OTU-based relative abundance profiles generated from 16S rRNA gene sequences are effective feature sets for classifying stool samples as originating from individuals with healthy, adenomatous, or cancerous colons [@Zackular:2014fba; @Baxter:2016dja]. The exceptional performance of bacteria in these classification models supports a role for bacteria in colorectal cancer. We built off of these findings by evaluating the ability of virus community signatures to classify stool samples and compared their performance to models built using bacterial community signatures.

@@

## Shifted Phage Influence Between Cancer Progression Stages
Because previous work has identified shifts in which bacteria were most important at different stages of colorectal cancer [@Zackular:2014fba; @Zackular:2016en; @Flynn:2016iu], we explored whether shifts in the relative influence of specific phages could be detected between healthy, adenomatous, and cancerous colons. We evaluated community shifts between the two disease stage transitions (healthy to adenomatous and adenomatous to cancerous) by building random forest models to compare only the diagnosis groups around the transitions. While bacterial OTU models performed equally well for all disease class comparisons, the virome model performances differed **(Figure \ref{transitionmodels} A-B)**. Like bacteria **(Figure \ref{transitionmodels} F-H)**, different virome members were important between the healthy to adenomatous and adenomatous to cancerous stages **(Figure \ref{transitionmodels} C-E)**.

@@

## Bacteriophage Dominance in Colorectal Cancer Virome
Differences in the colorectal cancer virome could have been driven directly by eukaryotic viruses or indirectly by bacteriophages acting through their bacterial hosts. To better understand the types of viruses that were important for colorectal cancer, we identified the virome OVUs as being similar to either eukaryotic viruses or bacteriophages. The most important viruses to the classification model were identified as bacteriophages (**Figure \ref{threewaymodel})**. Overall, we were able to identify `r signif(idcount, digits = sigfig)`% of the OVUs as known viruses, and `r signif(phagecount, digits = sigfig)`% of those viral OVUs aligned to bacteriophage reference genomes. It is important to note that this could have been influenced by our methodological biases against enveloped viruses (more common of eukaryotic viruses than bacteriophage), due to chloroform and DNase treatment for purification.

@@

## Working Model for Virome & Cancer Progression
Because of their propensity for mutagenesis and capacity for modulating their host functionality, many viruses are oncogenic [@Feng:2008kr; @Shuda:2011gf; @Schiller:2012ba; @Chang:1994up]. Some bacteria also have oncogenic properties, suggesting that bacteriophages may play an indirect role in promoting carcinogenesis by influencing bacterial community composition and dynamics [@Zackular:2014fba; @Garrett:2015fg; @Baxter:2014hb]. Despite their carcinogenic potential and the strong association between bacteria and colorectal cancer, a mechanistic link between virus colorectal communities and colorectal cancer has yet to be evaluated. Here we show that, like colonic bacterial communities, the colon virome was altered in patients with colorectal cancer relative to those with healthy colons. Our findings support a working hypothesis for oncogenesis by phage-modulated bacterial community composition.

@@

@@

https://github.com/SchlossLab/Hannigan_CRCVirome_Science_2017

>>>>

---
title: Diagnostics & Interactive Dynamics of the Colorectal Cancer Virome
author:
- name: Geoffrey D Hannigan
  affiliation: 1
- name: Melissa B Duhaime
  affiliation: 2
- name: Mack T Ruffin IV
  affiliation: 3
- name: Charlie C Koumpouras
  affiliation: 1
- name: Patrick D Schloss
  affiliation: 1,\*
address:
  - code: 1
    address: Department of Microbiology & Immunology, University of Michigan, Ann Arbor, Michigan, 48109
  - code: 2
    address: Department of Ecology & Evolutionary Biology, University of Michigan, Ann Arbor, Michigan, 48109
  - code: 3
    address: Department of Family & Community Medicine, Pennsylvania State University Hershey Medical Center, Hershey, Pennsylvania, 17033
  - code: \*
    address: To whom correspondence may be addressed.
output:
  md_document:
    variant: markdown
abstract: Viruses are assocaited with many human cancers, largely due to their mutagenic and functionally manipulative abilities. Despite this, cancer microbiome studies have almost exclusively focused on bacteria instead of viruses. We began evaluating the cancer virome by focusing on colorectal cancer, a primary cause of morbidity and mortality throughout the world, and a cancer linked to altered colonic bacterial community compositions while the virome role remains unknown. We used 16S rRNA gene, whole shotgun metagenomic, and purified virus metagenomic sequencing of stool to evaluate the differences in human colorectal cancer virus and bacterial community composition. Through random forest modeling we identified differences in the healthy and colorectal cancer virome. The cancer-associated virome consisted primarily of temperate bacteriophages that were also bacteria-virus community network hubs. These results provide foundational evidence that bacteriophage communities are associated with colorectal cancer and likely impact cancer progression by altering the bacterial host communities.
font: Times
fontsize: 12pt
---

@@

## Sample Collection and Processing
Our study cohort consisted of 90 human subjects, 30 of whom had healthy colons, 30 of whom had adenomas, and 30 of whom had carcinomas **(Figure \ref{sampleproc})**. Half of each stool sample was used to sequence the bacterial communities using both 16S rRNA gene and shotgun sequencing techniques. The 16S rRNA gene sequencing was performed for a previous study, and the sequences were re-analyzed using contemporary methods [@Zackular:2014fba]. The other half of each stool sample was purified for virus like particles (VLPs) before genomic DNA extraction and shotgun metagenomic sequencing. In the VLP purification, cells were disrupted and extracellular DNA degraded **(Figure \ref{sampleproc})** to allow the exclusive analysis of viral DNA within virus capsids. In this manner, the *extracellular virome* of encapsulated viruses was targeted.

@@

## Unaltered Diversity in Colorectal Cancer
Microbiome and disease associations are often described as being of an altered diversity (i.e., "dysbiotic"). Therefore, we first evaluated the influence of colorectal cancer on virome OVU diversity. We evaluated differences in communities between disease states using the Shannon diversity, richness, and Bray-Curtis metrics. We observed no significant alterations in either Shannon diversity or richness in the diseased states as compared to the healthy state **(Figure \ref{betaogu} C-D)**. There was no statistically significant clustering of the disease groups (ANOSIM p-value `r pvalcalc(alphadf[c(alphadf$Title %in% "anovawithoutneg"),"Stat"])`, **Figure \ref{betaogu}**). Notably, there was a significant difference between the few blank controls that remained after rarefying the data and the other study groups (ANOSIM p-value `r pvalcalc(alphadf[c(alphadf$Title %in% "anovawithneg"),"Stat"])`, **Figure \ref{betaogunegative})**, further supporting the quality of the sample set. In summary, standard alpha and beta diversity metrics were insufficient for capturing virus community differences between disease states **(Figure \ref{betaogu})**. This is consistent with what has been observed when the same metrics were applied to 16S rRNA sequenced and metagenomic samples [@Zeller:2014ix; @Zackular:2014fba; @Baxter:2016dja] and points to the need for alternate approaches to detect the impact of colorectal cancer disease state on these communities.

## Virome Composition in Colorectal Cancer
As opposed to the diversity metrics discussed above, OTU-based relative abundance profiles generated from 16S rRNA gene sequences are effective feature sets for classifying stool samples as originating from individuals with healthy, adenomatous, or cancerous colons [@Zackular:2014fba; @Baxter:2016dja]. The exceptional performance of bacteria in these classification models supports a role for bacteria in colorectal cancer. We built off of these findings by evaluating the ability of virus community signatures to classify stool samples and compared their performance to models built using bacterial community signatures.

@@

## Phage Influence Between CRC Stages
Because previous work has identified shifts in which bacteria were most important at different stages of colorectal cancer [@Zackular:2014fba; @Zackular:2016en; @Flynn:2016iu], we explored whether shifts in the relative influence of specific phages could be detected between healthy, adenomatous, and cancerous colons. We evaluated community shifts between the two disease stage transitions (healthy to adenomatous and adenomatous to cancerous) by building random forest models to compare only the diagnosis groups around the transitions. While bacterial OTU models performed equally well for all disease class comparisons, the virome model performances differed **(Figure \ref{transitionmodels} A-B)**. Like bacteria **(Figure \ref{transitionmodels} F-H)**, different virome members were important between the healthy to adenomatous and adenomatous to cancerous stages **(Figure \ref{transitionmodels} C-E)**.

@@

## Phage Dominance in CRC Virome
Differences in the colorectal cancer virome could have been driven directly by eukaryotic viruses or indirectly by bacteriophages acting through their bacterial hosts. To better understand the types of viruses that were important for colorectal cancer, we identified the virome OVUs as being similar to either eukaryotic viruses or bacteriophages. The most important viruses to the classification model were identified as bacteriophages (**Figure \ref{threewaymodel})**. Overall, we were able to identify `r signif(idcount, digits = sigfig)`% of the OVUs as known viruses, and `r signif(phagecount, digits = sigfig)`% of those viral OVUs aligned to bacteriophage reference genomes. It is important to note that this could have been influenced by our methodological biases against enveloped viruses (more common of eukaryotic viruses than bacteriophage), due to chloroform and DNase treatment for purification.

@@

## Model for Virome & Cancer Progression
Because of their propensity for mutagenesis and capacity for modulating their host functionality, many viruses are oncogenic [@Feng:2008kr; @Shuda:2011gf; @Schiller:2012ba; @Chang:1994up]. Some bacteria also have oncogenic properties, suggesting that bacteriophages may play an indirect role in promoting carcinogenesis by influencing bacterial community composition and dynamics [@Zackular:2014fba; @Garrett:2015fg; @Baxter:2014hb]. Despite their carcinogenic potential and the strong association between bacteria and colorectal cancer, a mechanistic link between virus colorectal communities and colorectal cancer has yet to be evaluated. Here we show that, like colonic bacterial communities, the colon virome was altered in patients with colorectal cancer relative to those with healthy colons. Our findings support a working hypothesis for oncogenesis by phage-modulated bacterial community composition.

@@

## Materials and Methods
This study was approved by the University of Michigan Institutional Review Board and all subjects provided informed consent. All study sequences are available on the NCBI Sequence Read Archive under the BioProject ID `PRJNA389927`. All associated source code is available at the following GitHub repository. Further detailed methods can be found in the supplemental methods.

https://github.com/SchlossLab/Hannigan_CRCVirome_Nature_2017

@@

https://github.com/SchlossLab/Hannigan_CRCVirome_Nature_2017
